Inference of proto-neutron star properties from gravitational-wave data in core-collapse supernovae

### 2020-11-20 ccsn.tex
cleanup biblio

\usepackage{amsmath}
\usepackage{color}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{wasysym}
\usepackage{psfrag}
%\usepackage{comment}
\usepackage{ifthen}
\usepackage[citecolor=blue,colorlinks=true]{hyperref}
\usepackage{longtable}
\usepackage{float}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{units}
\usepackage[title]{appendix}
\usepackage{aas_macros}
\usepackage{ulem}
\usepackage{subfigure}
\usepackage{multirow}
%\newcommand{\note}[1]{{\color{red}#1}}
\newcommand{\second}{\ensuremath{\mathrm{\,s}}}
%\newcommand{\mnras}{MNRAS}
\def\apjl{Astrophys. J. Lett.}
\def\prd{Phys.~Rev.~D}
\def\aap{A\&A}

@@

\begin{abstract}
The eventual detection of gravitational waves from core-collapse supernovae (CCSN) will help improve our current understanding of the explosion mechanism of massive stars. The stochastic nature of the late post-bounce gravitational wave signal due to the non-linear dynamics of the matter involved and the large number of degrees of freedom of the phenomenon make the source parameter inference problem very challenging. In this paper we take a step towards that goal and present a parameter estimation approach which is based on the gravitational waves associated with oscillations of proto-neutron stars (PNS). Numerical simulations of CCSN have shown that buoyancy-driven g-modes are responsible for a significant fraction of the gravitational wave signal and their time-frequency evolution is linked to the physical properties of the compact remnant through universal relations, as demonstrated in~\cite{Torres:2019b}. We use a set of 1D CCSN simulations to build a model that relates the  evolution of the PNS properties with the frequency of the dominant g-mode, which is extracted from the gravitational-wave data using a new algorithm we have developed for our study. The model is used to infer the time evolution of a combination of the mass and the radius of the PNS. The performance of the method is estimated employing simulations of 2D CCSN waveforms covering a progenitor mass range between 11 and 40 solar masses and different equations of state. Considering signals embedded in Gaussian gravitational wave detector noise, we show that it is possible to infer PNS properties for a galactic source using Advanced LIGO and Advanced Virgo data at design sensitivities. Third generation detectors such as Einstein Telescope and Cosmic Explorer will allow to test distances of ${\cal O}(100\, {\rm kpc})$. %\tf{Caveats? Gaussian noise?}
% 
%  Core-collapse supernovae are very important phenomena in the Universe and their expected gravitational
%  waves emission is a promising tool to study the onset of the supernova explosion mechanism. The complexity
%  of the gravitational wave signal due to matter effects and the large degrees of freedom of the phenomena
%  makes the source parameter inference problem very challenging. In this paper, we have considered the
%  proto neutron star oscillation modes that carry out a large fraction of the gravitational wave signal and
%  whose frequency time evolution is linked to the system properties through universal relations as
%  demonstrated in ~\cite{Torres:2019b}. We have developed a simple algorithm to extract from the
%  gravitational wave data the frequency evolution of the gravity mode. A set of 1D core collapse
%  supernova simulations is used to build a model of evolution of the proto neutron star properties
%  with the gravity mode frequency. The model is then used to infer the time evolution of a combinaison of
%  the radius and mass of the proto neutron star. We have then estimated the performance of the method by
%  performing simulations with 2D core collapse supernova waveforms covering a progenitor mass range between
%  11 and 40 solar mass and different equations of state. We have shown that for Advanced LIGO and Advanced
%  Virgo detectors at design sensitivity it will be possible to infer proton neutron star properties for
%  a galactic source, while third generation detectors Einstein Telescope and Cosmic Explorer will allow
%  to test distances of several hundreds of kiloparsec.

>>>>

\usepackage{amsmath}
\usepackage{color}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{wasysym}
\usepackage{psfrag}
%\usepackage{comment}
\usepackage{ifthen}
\usepackage[citecolor=blue,colorlinks=true]{hyperref}
\usepackage{longtable}
\usepackage{float}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{units}
\usepackage[title]{appendix}
\usepackage{aas_macros}
\usepackage{ulem}
\usepackage{subfigure}
\usepackage{multirow}
%\newcommand{\note}[1]{{\color{red}#1}}
\newcommand{\second}{\ensuremath{\mathrm{\,s}}}
%\newcommand{\mnras}{MNRAS}
\def\mnras{Mon. Not. Roy. Astron. Soc.}
\def\apjl{Astrophys. J. Lett.}
\def\apj{Astrophys. J.}
\def\prd{Phys.~Rev.~D}
\def\aap{A\&A}

@@

\begin{abstract}
The eventual detection of gravitational waves from core-collapse supernovae (CCSN) will help improve our current understanding of the explosion mechanism of massive stars. The stochastic nature of the late post-bounce gravitational wave signal due to the non-linear dynamics of the matter involved and the large number of degrees of freedom of the phenomenon make the source parameter inference problem very challenging. In this paper we take a step towards that goal and present a parameter estimation approach which is based on the gravitational waves associated with oscillations of proto-neutron stars (PNS). Numerical simulations of CCSN have shown that buoyancy-driven g-modes are responsible for a significant fraction of the gravitational wave signal and their time-frequency evolution is linked to the physical properties of the compact remnant through universal relations, as demonstrated in~\citep{Torres:2019b}. We use a set of 1D CCSN simulations to build a model that relates the  evolution of the PNS properties with the frequency of the dominant g-mode, which is extracted from the gravitational-wave data using a new algorithm we have developed for our study. The model is used to infer the time evolution of a combination of the mass and the radius of the PNS. The performance of the method is estimated employing simulations of 2D CCSN waveforms covering a progenitor mass range between 11 and 40 solar masses and different equations of state. Considering signals embedded in Gaussian gravitational wave detector noise, we show that it is possible to infer PNS properties for a galactic source using Advanced LIGO and Advanced Virgo data at design sensitivities. Third generation detectors such as Einstein Telescope and Cosmic Explorer will allow to test distances of ${\cal O}(100\, {\rm kpc})$. %\tf{Caveats? Gaussian noise?}
% 
%  Core-collapse supernovae are very important phenomena in the Universe and their expected gravitational
%  waves emission is a promising tool to study the onset of the supernova explosion mechanism. The complexity
%  of the gravitational wave signal due to matter effects and the large degrees of freedom of the phenomena
%  makes the source parameter inference problem very challenging. In this paper, we have considered the
%  proto neutron star oscillation modes that carry out a large fraction of the gravitational wave signal and
%  whose frequency time evolution is linked to the system properties through universal relations as
%  demonstrated in ~\cite{Torres:2019b}. We have developed a simple algorithm to extract from the
%  gravitational wave data the frequency evolution of the gravity mode. A set of 1D core collapse
%  supernova simulations is used to build a model of evolution of the proto neutron star properties
%  with the gravity mode frequency. The model is then used to infer the time evolution of a combinaison of
%  the radius and mass of the proto neutron star. We have then estimated the performance of the method by
%  performing simulations with 2D core collapse supernova waveforms covering a progenitor mass range between
%  11 and 40 solar mass and different equations of state. We have shown that for Advanced LIGO and Advanced
%  Virgo detectors at design sensitivity it will be possible to infer proton neutron star properties for
%  a galactic source, while third generation detectors Einstein Telescope and Cosmic Explorer will allow
%  to test distances of several hundreds of kiloparsec.

### 2020-11-20 conclusion.tex
cleanup biblio

The detection of GW from CCSN may help improve our current understanding of the explosion mechanism of massive stars.  
In this paper we have proposed an exploratory method to infer PNS properties using an approach based on GW associated with convective oscillations of PNS. As shown by~\cite{Torres:2019b} buoyancy-driven g-modes are excited in numerical simulations of CCSN and their time-frequency evolution is linked to the physical properties of the compact remnant through universal relations. Such modes are responsible for a significant fraction of the highly stochastic GW emitted after core bounce. The findings reported in this paper suggest that PNS asteroseismology might be within reach of current and third-generation GW detectors.

In our study we have used a set of 1D CCSN simulations to build a model that relates the  evolution of PNS properties with the frequency of the dominant g-mode, namely the $\mbox{}^2g_2$ mode. This relationship is extracted from the GW data using an algorithm developed for this investigation. This algorithm is  a first attempt to infer the time evolution of a particular combination of the PNS mass and radius based on the universal relations found in~\cite{Torres:2019b}. More precisely, we have considered the ratio $r=M_{\rm PNS}/R_{\rm PNS}^2$ (the PNS surface gravity) derived from the observation of the $\mbox{}^2g_2$ oscillation mode in the numerically generated GW data. The  performance of our method has been estimated employing  simulations of 2D CCSN waveforms covering a progenitor mass range between 11 and 40 solar masses and different equations of state.

>>>>

The detection of GW from CCSN may help improve our current understanding of the explosion mechanism of massive stars.  
In this paper we have proposed an exploratory method to infer PNS properties using an approach based on GW associated with convective oscillations of PNS. As shown by~\citep{Torres:2019b} buoyancy-driven g-modes are excited in numerical simulations of CCSN and their time-frequency evolution is linked to the physical properties of the compact remnant through universal relations. Such modes are responsible for a significant fraction of the highly stochastic GW emitted after core bounce. The findings reported in this paper suggest that PNS asteroseismology might be within reach of current and third-generation GW detectors.

In our study we have used a set of 1D CCSN simulations to build a model that relates the  evolution of PNS properties with the frequency of the dominant g-mode, namely the $\mbox{}^2g_2$ mode. This relationship is extracted from the GW data using an algorithm developed for this investigation. This algorithm is  a first attempt to infer the time evolution of a particular combination of the PNS mass and radius based on the universal relations found in~\citep{Torres:2019b}. More precisely, we have considered the ratio $r=M_{\rm PNS}/R_{\rm PNS}^2$ (the PNS surface gravity) derived from the observation of the $\mbox{}^2g_2$ oscillation mode in the numerically generated GW data. The  performance of our method has been estimated employing  simulations of 2D CCSN waveforms covering a progenitor mass range between 11 and 40 solar masses and different equations of state.

### 2020-11-20 introduction.tex
cleanup biblio

The life of sufficiently massive stars, i.e.~those born with masses between $\sim 8$~M$_\odot$ and $\sim 120$~M$_\odot$, ends with the collapse of {the} iron core under {its} own gravity, leading {to} the formation of a neutron star {(NS)} or a black hole (BH), {and} followed (typically but not necessarily in the BH case) by {a supernova} explosion. Nearby core-collapse supernova (CCSN) explosions are expected to be sources of gravitational waves (GWs) and they could be 
%are one of the main candidates for 
the next great discovery of current ground-based observatories. However, these are relative rare events. A neutrino-driven explosion \citep{Bethe:1990} is the most likely outcome in the case of slowly rotating cores, which are present in the bulk of CCSN progenitors. The emitted GWs could be detected with the advanced ground-based GW detector network, Advanced LIGO (aLIGO)~\citep{TheLIGOScientific:2014jea}, Advanced Virgo (AdV)~\citep{TheVirgo:2014hva} and
KAGRA~\citep{Aso:2013eba}, within $\sim$ \unit[5]{kpc} \citep{Gossan:2016,TargetedSNSearchO12}. Such a galactic event has a rate of about $2-3$ per century \citep{Adams:2013,Rozwadowska:2021}.
For the case of rapidly rotating progenitor cores the result is likely a magneto-rotational explosion, yielding  a more powerful GW signal that could be detected within \unit[50]{kpc} and, for some extreme models, up to \unit[5--30]{Mpc} \citep{Gossan:2016,TargetedSNSearchO12}. However, only about $1\%$ of the electromagnetically observed events show signatures of fast rotation (broad-lined type Ic SNe \citep{Li:2011b} or events associated with long GRBs \citep{Chapman:2007}) making this possibility a subdominant channel of detection with a galactic event rate of $\sim$ \unit[$10^{-4}$]{$\rm {yr}^{-1}$}.
%For the results discussed in this work we only consider neutrino-driven CCSN of non-rotating stellar cores.%\tf{This last sentence must be mentioned somewhere (or maybe not) but perhaps here is not the best place.}\mab{MAB:I'm in favor of deleting it as it is rather obvious we are considering neutrino driven CCSN}
Despite the low rates, CCSN are of great scientific interest because they produce complex GW signals which could provide significant clues about the physical processes at work after the gravitational collapse of stellar cores.

In the last decade significant progress has been made in the development of numerical codes, {in particular in the treatment of multidimensioal effects \citep{BMueller:2020}.} In the case of  neutrino-driven explosions, the GW emission is {primarly induced by instabilities developed at the newly formed proto-neutron star (PNS) and by the non-spherical accreting flow of hot matter over its surface \citep{Kotake:2017}.  These} dynamics excite the different modes of oscillation of the PNS which ultimately leads to the emission of GWs. The frequency and time evolution of these modes carry information about the properties of the GW emitter and could allow to perform PNS asteroseismology.

@@

All multidimensional numerical simulations show the systematic appearance in time-frequency diagrams (or spectrograms) of a distinct and relatively narrow feature 
%
%{The main feature appearing systematically in the GW spectrum of multidimensional numerical simulations is a strong and relatively narrow \mab{oscillation} 
%
during the post-bounce evolution of the system, with frequency rising 
from about \unit[100]{Hz} up to a few kHz (at most) and a typical duration of \unit[0.5 -- 1]{s}. This feature has been interpreted as a continuously excited gravity mode (g-mode, see \citep{kokkotas,Friedman:2013} for a definition in this context) of the PNS \citep{murphy:09, mueller:13gw, Cerda:2013, Yakunin:2015, Kuroda:2016, Andresen:2017}. 
In these models the monotonic raise of the frequency of the mode is related to the contraction of the PNS. The {typical} frequencies of {these} modes make them interesting targets for
%promising {source} for 
ground-based GW interferometers.

 The {properties of} g-modes in hot {PNSs} have been studied since the 1990s
 %end of last century 
 {by means of linear perturbation analysis of background PNS models}. The oscillation modes connected with the surface of hot PNSs were first considered by McDermott {\it et al.} \citep{McDermott:1983}. Additionally, the stratified structure of the PNS allows for the presence of different types of g-modes related to the fluid core \citep{Reisenegger:1992}. Many subsequent works used simplified neutron star models assuming an equilibrium configuration {as a background}, to study the effect of rotation \citep{Ferrari:2004}, general relativity \citep{Passamonti:2005}, non-linearities \citep{Dimmelmeier:2006}, phase transitions \citep{Kruger:2015} and realistic equation of state \citep{Camelio:2017}. {Only recently, there have been efforts to incorporate more suitable backgrounds based on numerical simulations in the computation of the mode structure and evolution \citep{Sotani:2016,Torres:2018, Morozova:2018, Torres:2019a,Torres:2019b,Sotani:2019,WS:2019,Sotani:2020a, Sotani:2020b}}.

Using results from 2D CCSN numerical simulations as a background~\citep{Torres:2018, Torres:2019a} found that the eigenmode spectrum of the region within the shock (including the PNS and the post-shock region) 
shows a good match to the mode frequencies and to the features observed in the GW spectrum of the same simulations (specially when space-time perturbations are included \citep{Torres:2019a}).
%\tf{This last sentence might be improved.} \mab{MAB: What about: Computing the mode spectrum of the region within the shock (including the PNS and the post-shock region) consists in solving an eigenvalue problem using background simulations. Using results from 2D CCSN numerical simulations \citep{Torres:2018, Torres:2019a} found that the eigenmode spectrum shows a good match to the mode frequencies and to the features observed in the GW spectrum of the same simulations (specially when space-time perturbations are included \citep{Torres:2019a}). } \pcd{[PCD: I am not convinced about the first sentence of mab's proposal. Maybe if we just delete "with respect to which to solve an eigenvalue problem" the sentence becomes more readable]}

This reveals that it is posible to perform CCSN asteroseismology {under realistic conditions} and serves as a starting point to carry out inference of astrophysical parameters of PNSs. 
Further work was presented in {\citep{Torres:2019b} who found that it is possible to derive simple relations} between the instantaneous frequency of the g-mode and the mass and radius of the PNS {at each time of the numerical evolutions}. These relations are universal as they do not depend {on the equation of state (EOS) or on the mass of the progenitor {and they only depend weakly on} the numerical code used {(see discussion in Section~\ref{sec:simulations})}. {Similar universal relations have been discussed by \citep{Sotani:2020a,Sotani:2020b} who also found that they do not depend on the dimensionality (1D, 2D or 3D) of the numerical simulation used as a background.

{In this work we introduce a method to infer PNS properties, namely a combination of the mass and radius, using GW information. For this purpose we have developed an algorithm to  
extract the time-frequency evolution of the main feature in the spectrograms of the GW emission of 2D simulations of CCSN. This feature corresponds to the $^2\rm{g}_2$ mode, according to the nomenclature used in \citep{Torres:2019b} (different authors may have slightly different naming convention). Next, we use the universal relations obtained by \citep{Torres:2019b}{, based on a set of 1D simulations,} to infer the time evolution of the ratio $M_{\rm PNS}/R_{\rm PNS}^2$ (the PNS surface gravity), where  $M_{\rm PNS}$ and $R_{\rm PNS}$ are the mass and the radius of the PNS.} Finally, using 2D CCSN waveform corresponding to different progenitor masses we estimate the performance of the algorithm for current and future generation of ground-based GW detectors.

>>>>

The life of sufficiently massive stars, i.e.~those born with masses between $\sim 8$~M$_\odot$ and $\sim 120$~M$_\odot$, ends with the collapse of {the} iron core under {its} own gravity, leading {to} the formation of a neutron star {(NS)} or a black hole (BH), {and} followed (typically but not necessarily in the BH case) by {a supernova} explosion.
Nearby core-collapse supernova (CCSN) explosions are expected to be sources of gravitational waves (GWs) and they could be the next great discovery of current ground-based observatories. However, these are relative rare events. A neutrino-driven explosion~\citep{Bethe:1990} is the most likely outcome in the case of slowly rotating cores, which are present in the bulk of CCSN progenitors.
The emitted GWs could be detected with the advanced ground-based GW detector network, Advanced LIGO (aLIGO)~\citep{TheLIGOScientific:2014jea}, Advanced Virgo (AdV)~\citep{TheVirgo:2014hva} and KAGRA~\citep{Aso:2013eba}, within $\sim$ \unit[5]{kpc}~\citep{Gossan:2016,TargetedSNSearchO12}. Such a galactic event has a rate of about $2-3$ per century~\citep{Adams:2013,Rozwadowska:2021}.
For the case of rapidly rotating progenitor cores the result is likely a magneto-rotational explosion, yielding  a more powerful GW signal that could be detected within \unit[50]{kpc} and, for some extreme models, up to \unit[5--30]{Mpc}~\citep{Gossan:2016,TargetedSNSearchO12}. However, only about $1\%$ of the electromagnetically observed events show signatures of fast rotation (broad-lined type Ic SNe~\citep{Li:2011b} or events associated with long GRBs~\citep{Chapman:2007}) making this possibility a subdominant channel of detection with a galactic event rate of $\sim$ \unit[$10^{-4}$]{$\rm {yr}^{-1}$}.
Despite the low rates, CCSN are of great scientific interest because they produce complex GW signals which could provide significant clues about the physical processes at work after the gravitational collapse of stellar cores.

In the last decade significant progress has been made in the development of numerical codes, {in particular in the treatment of multidimensional effects~\citep{BMueller:2020}.} In the case of  neutrino-driven explosions, the GW emission is {primarly induced by instabilities developed at the newly formed proto-neutron star (PNS) and by the non-spherical accreting flow of hot matter over its surface~\citep{Kotake:2017}.  These} dynamics excite the different modes of oscillation of the PNS which ultimately leads to the emission of GWs. The frequency and time evolution of these modes carry information about the properties of the GW emitter and could allow to perform PNS asteroseismology.

@@

All multidimensional numerical simulations show the systematic appearance in time-frequency diagrams (or spectrograms) of a distinct and relatively narrow feature 
%
%{The main feature appearing systematically in the GW spectrum of multidimensional numerical simulations is a strong and relatively narrow \mab{oscillation} 
%
during the post-bounce evolution of the system, with frequency rising 
from about \unit[100]{Hz} up to a few kHz (at most) and a typical duration of \unit[0.5 -- 1]{s}. This feature has been interpreted as a continuously excited gravity mode (g-mode, see~\citep{kokkotas,Friedman:2013} for a definition in this context) of the PNS~\citep{murphy:09, mueller:13gw, Cerda:2013, Yakunin:2015wra, Kuroda:2016, Andresen:2017}. 
In these models the monotonic raise of the frequency of the mode is related to the contraction of the PNS. The {typical} frequencies of {these} modes make them interesting targets for
%promising {source} for 
ground-based GW interferometers.

 The {properties of} g-modes in hot {PNSs} have been studied since the 1990s
 %end of last century 
 {by means of linear perturbation analysis of background PNS models}. The oscillation modes connected with the surface of hot PNSs were first considered by McDermott {\it et al.}~\citep{McDermott:1983}. Additionally, the stratified structure of the PNS allows for the presence of different types of g-modes related to the fluid core~\citep{Reisenegger:1992}. Many subsequent works used simplified neutron star models assuming an equilibrium configuration {as a background}, to study the effect of rotation~\citep{Ferrari:2004}, general relativity \citep{Passamonti:2005}, non-linearities~\citep{Dimmelmeier:2006}, phase transitions~\citep{Kruger:2015} and realistic equation of state~\citep{Camelio:2017}. {Only recently, there have been efforts to incorporate more suitable backgrounds based on numerical simulations in the computation of the mode structure and evolution~\citep{Sotani:2016,Torres:2018, Morozova:2018, Torres:2019a,Torres:2019b,Sotani:2019,WS:2019,Sotani:2020a, Sotani:2020b}}.

Using results from 2D CCSN numerical simulations as a background~\citep{Torres:2018, Torres:2019a} found that the eigenmode spectrum of the region within the shock (including the PNS and the post-shock region) 
shows a good match to the mode frequencies and to the features observed in the GW spectrum of the same simulations (specially when space-time perturbations are included~\citep{Torres:2019a}).
%\tf{This last sentence might be improved.} \mab{MAB: What about: Computing the mode spectrum of the region within the shock (including the PNS and the post-shock region) consists in solving an eigenvalue problem using background simulations. Using results from 2D CCSN numerical simulations \citep{Torres:2018, Torres:2019a} found that the eigenmode spectrum shows a good match to the mode frequencies and to the features observed in the GW spectrum of the same simulations (specially when space-time perturbations are included \citep{Torres:2019a}). } \pcd{[PCD: I am not convinced about the first sentence of mab's proposal. Maybe if we just delete "with respect to which to solve an eigenvalue problem" the sentence becomes more readable]}

This reveals that it is posible to perform CCSN asteroseismology {under realistic conditions} and serves as a starting point to carry out inference of astrophysical parameters of PNSs. 
Further work was presented in~\citep{Torres:2019b} who found that it is possible to derive simple relations between the instantaneous frequency of the g-mode and the mass and radius of the PNS {at each time of the numerical evolutions}. These relations are universal as they do not depend {on the equation of state (EOS) or on the mass of the progenitor {and they only depend weakly on} the numerical code used {(see discussion in Section~\ref{sec:simulations})}. {Similar universal relations have been discussed by~\citep{Sotani:2020a,Sotani:2020b} who also found that they do not depend on the dimensionality (1D, 2D or 3D) of the numerical simulation used as a background.

Many previous data analysis efforts focused on the reconstruction of the gravitational wave strain amplitude
without assuming a particular signal model~\citep{Summerscales:2008,Klimenko:2015ypf,CornLitt}. As numerical
simulations became more available other methods using catalogs of waveforms have been proposed to identify
the supernova explosion mechanism.
Among them, principal component analysis helps at reducing the complexity of CCSN waveform catalogs to fewer
parameters~\citep{Heng:2009,roever:09,Edwards:2014,powell:2016,powell:2017,Roma:2019kcd}.
The classification challenge is also well addressed with deep learning methods~\citep{astone:2018,Chan:2019fuz}.

{In this work we introduce a method to infer PNS properties, namely a combination of the mass and radius, using GW information. For this purpose we have developed an algorithm to  
extract the time-frequency evolution of the main feature in the spectrograms of the GW emission of 2D simulations of CCSN. This feature corresponds to the $^2\rm{g}_2$ mode, according to the nomenclature used in~\citep{Torres:2019b} (different authors may have slightly different naming convention). Next, we use the universal relations obtained by~\citep{Torres:2019b}{, based on a set of 1D simulations,} to infer the time evolution of the ratio $M_{\rm PNS}/R_{\rm PNS}^2$ (the PNS surface gravity), where  $M_{\rm PNS}$ and $R_{\rm PNS}$ are the mass and the radius of the PNS.} Finally, using 2D CCSN waveform corresponding to different progenitor masses we estimate the performance of the algorithm for current and future generation of ground-based GW detectors.

### 2020-11-20 methods.tex
cleanup biblio

We next outline our strategy for estimating the time evolution of $r(t)$
%ratio $r=M_{\rm PNS}/R_{\rm PNS}^2$ (in units of solar mass and km) 
from the observation of the $\mbox{}^2g_2$ oscillation mode in the GW detector data.
%An integral part of this strategy is the universal relations that relate the
%characteristic frequency of the PNS oscillation $f$, $g$ and $p$ modes with the mass
%and the radius of the PNS, the shock radius and the total mass inside the shock as
%demonstrated in \cite{Torres:2019b}.
To build the model of the ratio $r$ as a function of the frequency $f$ we use the 
1D simulations of the {\it model set}. Figure~\ref{fig:LMVAR}
shows the data for the $18$ numerical simulations. 
%As identified by \cite{Torres:2019b}, the only systematic deviation from a single universal relation is the numerical code used in the simulations. 
%To avoid any systematic effect, we only use the $18$ simulations performed with the {\sc AENUS-ALCAR}
%code, which is the same code that was used in our test set.
%\mab{We defer a discussion of the possible consequences of this choice in Section \ref{sec:conclusion}.}
%The consequences of this choice are discussed in the conclusions.
Using these data we parametrize the discretized ratio $r_i$ with a cubic polynomial
regression with heteroscedastic errors
\begin{equation}
\label{eq:model1}
r_i=\beta_1 f_i + \beta_2 f_i^2 +\beta_3 f_i^3 + \epsilon_i\,,
\end{equation}
where $\epsilon_i$ are assumed to be independent zero-mean Gaussian errors with
variances $\sigma_i^2$ that increase with frequency $f_i$. The model for frequency-dependent
variances is
\begin{equation}
\log \sigma_i=\alpha_0+ \alpha_1 f_i + \alpha_2 f_i^2 + \delta_i\,,
\end{equation}
with independent and identically zero-mean Gaussian errors $\delta_i$. The R-package \texttt{lmvar}
\cite{lmvar:2019} that implements a maximum likelihood approach was used to fit the model.

@@

{We use this model to infer the properties of the simulations in the 
{\it test set} discussed in Section \ref{sec:simulations}.}
To describe the method we focus on the GW signal
of model {\texttt s20S}, originally
sampled at \unit[16384]{Hz} but downsampled at \unit[4096]{Hz}.
A spectrogram of this signal is shown in Figure \ref{fig:spectrogram} based on
autoregressive estimates~\cite{BrockwellPeterJ1991Tsta} of the local spectra for successive time intervals of 
length 200 with a 90\% overlap.
The dominant emission mode corresponds to the PNS oscillation $\mbox{}^2 g_2$-mode. We have
developed a time-frequency method to track the ridge $m(t) $ in the spectrogram,
taking into account that it is monotonically increasing with time. 
This is a property of the $\mbox{}^2 g_2$-mode whose frequency   
increases as the PNS becomes more massive and compact.
Starting from either the left- or right-most column of the time-frequency matrix
we identify and trace the sequence of amplitude peaks within a certain frequency
band given the monotonicity constraint. Specific details about the reconstruction of the $\mbox{}^2 g_2$-mode ridge 
are provided in Appendix \ref{app:gmode}.

>>>>

We next outline our strategy for estimating the time evolution of $r(t)$
%ratio $r=M_{\rm PNS}/R_{\rm PNS}^2$ (in units of solar mass and km) 
from the observation of the $\mbox{}^2g_2$ oscillation mode in the GW detector data.
%An integral part of this strategy is the universal relations that relate the
%characteristic frequency of the PNS oscillation $f$, $g$ and $p$ modes with the mass
%and the radius of the PNS, the shock radius and the total mass inside the shock as
%demonstrated in \cite{Torres:2019b}.
To build the model of the ratio $r$ as a function of the frequency $f$ we use the 
1D simulations of the {\it model set}. Figure~\ref{fig:LMVAR}
shows the data for the $18$ numerical simulations. 
%As identified by \cite{Torres:2019b}, the only systematic deviation from a single universal relation is the numerical code used in the simulations. 
%To avoid any systematic effect, we only use the $18$ simulations performed with the {\sc AENUS-ALCAR}
%code, which is the same code that was used in our test set.
%\mab{We defer a discussion of the possible consequences of this choice in Section \ref{sec:conclusion}.}
%The consequences of this choice are discussed in the conclusions.
Using these data we parametrize the discretized ratio $r_i$ with a cubic polynomial
regression with heteroscedastic errors
\begin{equation}
\label{eq:model1}
r_i=\beta_1 f_i + \beta_2 f_i^2 +\beta_3 f_i^3 + \epsilon_i\,,
\end{equation}
where $\epsilon_i$ are assumed to be independent zero-mean Gaussian errors with
variances $\sigma_i^2$ that increase with frequency $f_i$. The model for frequency-dependent
variances is
\begin{equation}
\log \sigma_i=\alpha_0+ \alpha_1 f_i + \alpha_2 f_i^2 + \delta_i\,,
\end{equation}
with independent and identically zero-mean Gaussian errors $\delta_i$. The R-package \texttt{lmvar}~\citep{lmvar:2019} that implements a maximum likelihood approach was used to fit the model.

@@

{We use this model to infer the properties of the simulations in the 
{\it test set} discussed in Section \ref{sec:simulations}.}
To describe the method we focus on the GW signal
of model {\texttt s20S}, originally
sampled at \unit[16384]{Hz} but downsampled at \unit[4096]{Hz}.
A spectrogram of this signal is shown in Figure \ref{fig:spectrogram} based on
autoregressive estimates~\citep{BrockwellPeterJ1991Tsta} of the local spectra for successive time intervals of length 200 with a 90\% overlap.
The dominant emission mode corresponds to the PNS oscillation $\mbox{}^2 g_2$-mode. We have
developed a time-frequency method to track the ridge $m(t) $ in the spectrogram,
taking into account that it is monotonically increasing with time. 
This is a property of the $\mbox{}^2 g_2$-mode whose frequency   
increases as the PNS becomes more massive and compact.
Starting from either the left- or right-most column of the time-frequency matrix
we identify and trace the sequence of amplitude peaks within a certain frequency
band given the monotonicity constraint. Specific details about the reconstruction of the $\mbox{}^2 g_2$-mode ridge 
are provided in Appendix~\ref{app:gmode}.

### 2020-11-20 results.tex
cleanup biblio

To estimate how accurately we can infer the time evolution of $r(t)$ in the GW data of a single
detector we inject the GW signal for model {\texttt s20S} into 
100 Gaussian noise realisations whose power spectral density (PSD) follows the aLIGO
spectrum~\cite{aLIGOsens:2018}. %The PSD of aLIGO and AdV at design sensitivities, along with those of planned third-generation detectors, are shown in Figure~\ref{fig:spectrum}.

@@

We perform the same analysis using the design sensitivity curve of AdV and expected sensitivity curves for third-generation 
GW detectors. The results for the former are reported in Table \ref{tab:results} and are discussed below. We focus now on third-generation detectors, presenting our findings in Table \ref{tab:results} and in Fig.~\ref{fig:s20--SFHo_all3G}. In Europe the Einstein Telescope project proposes to host in a 10-km equilateral triangle configuration three low-power, low-frequency, cryogenic interferometers as well as three high-power, high-frequency interferometers. Three sensitivity curves, ET\_B, ET\_C and ET\_D corresponding to different options and stages of the project~\cite{Hild_2011} are considered in our study. %(cf.~Fig.~\ref{fig:spectrum}).
The US based project Cosmic Explorer~\cite{reitze2019cosmic} is expecting to reach its design
sensitivity circa 2040 through two phases labeled CE1 and CE2. %Their corresponding sensitivity curves are also shown in Fig.~\ref{fig:spectrum}.

>>>>

To estimate how accurately we can infer the time evolution of $r(t)$ in the GW data of a single
detector we inject the GW signal for model {\texttt s20S} into 
100 Gaussian noise realisations whose power spectral density (PSD) follows the aLIGO
spectrum~\citep{aLIGOsens:2018}. %The PSD of aLIGO and AdV at design sensitivities, along with those of planned third-generation detectors, are shown in Figure~\ref{fig:spectrum}.

@@

We perform the same analysis using the design sensitivity curve of AdV and expected sensitivity curves for third-generation 
GW detectors. The results for the former are reported in Table \ref{tab:results} and are discussed below. We focus now on third-generation detectors, presenting our findings in Table \ref{tab:results} and in Fig.~\ref{fig:s20--SFHo_all3G}. In Europe the Einstein Telescope project proposes to host in a 10-km equilateral triangle configuration three low-power, low-frequency, cryogenic interferometers as well as three high-power, high-frequency interferometers. Three sensitivity curves, ET\_B, ET\_C and ET\_D corresponding to different options and stages of the project~\citep{Hild_2011} are considered in our study. %(cf.~Fig.~\ref{fig:spectrum}).
The US based project Cosmic Explorer~\citep{reitze2019cosmic} is expecting to reach its design
sensitivity circa 2040 through two phases labeled CE1 and CE2. %Their corresponding sensitivity curves are also shown in Fig.~\ref{fig:spectrum}.
